Depression and suicidality as credible signals of need

### 2021-05-20 Paper.rmd
Fixed a few typos. Added edits in the form of commented out lines with the text "CurrentDraft" at the end. Added year to Clifford 2021 citation in MTBib.

There is strong evidence that most cases of depression are caused by major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, depression onsets soon after a negative event [@KendlerEtAl1999] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after; @rosenstrom2017]. Additional support for the causal role of negative life events comes from the fact that the relationship between negative events and depression holds true when looking at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Additionally, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably not due to mental illness but are instead functional instances of psychological pain [@Hagen:1999; @Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021].

@@

Formal models of bargaining highlight the role of private information regarding the value each party places the cooperative endeavor. Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to request an amount greater than their actual valuation, which would harm the other party. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of waiting for a better offer outweighs the low cost of delay whereas for those for those with a high valuation the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005]. @Hagen2003 proposed that the substantial reduction in productivity in depression corresponds to a willingness to delay, and is therefore a credible signal of low valuation and need. In other words, reduced productivity is relatively less costly for those whose efforts are currently not yielding many fitness-relevant benefits than for those whose efforts are yielding substantial benefits [@Hagen2003].

## Suicidality

@@

The final part of the survey was a brief demographic questionnaire which asked for (1) the age, (2) sex, (3) number of siblings, (4) number of sons, (5) number of daughters, (6) current relationship status, (7) highest level of education, and (8) the annual household income of each participant.

>>>>

<!-- Almost all individuals will suffer adversity, such as injury or loss of material or social resources, at some point in their lives. Psychological pain, such as sadness and low mood, probably evolved to motivate victims of adversity to shift their attention to the causes of the problem so as to mitigate its negative fitness consequences and to learn to avoid future such adverse events [@thornhill1989; @andrews2010; @Nesse1990]. In many types of adversity, social partners could help minimize the costs. Therefore signals of psychological pain, such as sad expressions and crying, probably evolved to indicate need [e.g., @balsters2013]. CurrentDraft-->

There is strong evidence that most cases of depression are caused by major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, depression onsets soon after a negative event [@KendlerEtAl1999] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after; @rosenstrom2017]. Additional support for the causal role of negative life events comes from the fact that the relationship between negative events and depression holds true when looking at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably not due to mental illness but are instead functional instances of psychological pain [@Hagen:1999; @Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021].

@@

Formal models of bargaining highlight the role of private information regarding the value each party places the cooperative endeavor. Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to request an amount greater than their actual valuation, which would harm the other party. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of waiting for a better offer outweighs the low cost of delay whereas for those with a high valuation the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005]. @Hagen2003 proposed that the substantial reduction in productivity in depression corresponds to a willingness to delay, and is therefore a credible signal of low valuation and need. In other words, reduced productivity is relatively less costly for those whose efforts are currently not yielding many fitness-relevant benefits than for those whose efforts are yielding substantial benefits [@Hagen2003].

<!-- Some evolutionary theorists have questioned the bargaining model based on evidence that depression is aversive to social partners [@nettle2004; @allen2006; @nesse2006; @myers2016]. These critics fail to acknowledge the central role of *conflict* in the signaling/bargaining models. Anger causes negative reactions in targeted social partners, yet it is probably an adaptation that exploits advantages in physical or social formidability to force concessions from others [@SellEtAl2009]. Aversive responses to bargaining are expected throughout the process, with potential helpers likely signaling the costliness of providing support in turn. Importantly, such responses do not prevent the delivery of meaningful benefits [@SymeHagen2021]. CurrentDraft [move this up from suicidality section-->

<!-- As cooperation generates benefits overtime, withholding or risking contributions incurs a cost of delay – the more one values the endeavor, the greater the cost of delay. CurrentDraft-->
## Suicidality

@@

The final part of the survey was a brief demographic questionnaire which asked for the (1) age, (2) sex, (3) number of siblings, (4) number of sons, (5) number of daughters, (6) current relationship status, (7) highest level of education, and (8) the annual household income of each participant.
